Equivalence Hypergraphs:\\ Semantics of E-Graphs and their Generalization to Monoidal Theories

### 2023-10-17 sections/introduction.tex

While equality saturation is a state-of-the-art optimization technique, the formal mathematical theory of e-graphs is perhaps underdeveloped. In this paper, we contribute a categorical understanding of e-graphs and their rewriting. Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories, where the semilattice \textit{join} allows us to consider formal sets of terms (string diagrams, morphisms).

>>>>

\update{Move focus from semantics of e-graphs to their extension to monoidal theories?}
While equality saturation is a state-of-the-art optimization technique, the formal mathematical theory of e-graphs is perhaps underdeveloped. In this paper, we contribute a categorical understanding of e-graphs and their rewriting. Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories, where the semilattice \textit{join} allows us to consider formal sets of terms (string diagrams, morphisms).
